\begin{thebibliography}{10}
\def\enquote#1{``#1''}
\expandafter\ifx\csname url\endcsname\relax
  \def\url#1{{\tt #1}}\fi
\expandafter\ifx\csname urlprefix\endcsname\relax\def\urlprefix{URL }\fi

\bibitem{baken1999}
Baken, R.~J. dan Orlikoff, R.~F.,
  {\em {Clinical Measurement of Speech \& Voice}\/},
  edisi ke-2, Singular, 1999.

\bibitem{benesty2008}
Benesty, J., Chen, J., dan Huang, Y.,
  {Direction-of-Arrival and Time-Difference-of-Arrival Estimation},
  {\em {Microphone Array Signal Processing}\/},
  181--215, Springer, 2008.

\bibitem{beun2003}
Beun, R., Vos, E.~D., dan Witteman, C.,
  {Embodied Conversational Agents: Effects on Memory Performance and Anthropomorphisation},
  {\em Intelligent Virtual Agents\/}, 
  315--319, Springer, 2003.

\bibitem{birchfield2005}
Birchfield, S. dan Gangishetty, R.,
  {Acoustic localization by interaural level difference},
  {\em Proceedings of IEEE International Conference
  on Acoustics, Speech, and Signal Processing (ICASSP'05)\/},
  \textbf{4}, IEEE, 2005.

\bibitem{cassell2000}
Cassell, J.,
  {Embodied conversational interface agents},
  {\em Communications of the ACM\/}, \textbf{43}(4),
  70--78, 2000.

\bibitem{cassell2000a}
Cassell, J., Bickmore, T., Vilhj{\'a}lmsson, H., dan Yan, H.,
  {More than just a pretty face: affordances of embodiment},
  {\em Proceedings of the 5th International Conference on Intelligent User Interfaces\/},
  52--59, ACM, 2000.

\bibitem{cassell1998}
Cassell, J., et~al.,
  {An architecture for embodied conversational characters},
  {\em Proceedings of the First Workshop on Embodied Conversational Characters\/},
  109--120, 1998.

\bibitem{cassell1999}
Cassell, J., et~al.,
  {Embodiment in conversational interfaces: Rea},
  {\em Proceedings of the SIGCHI Conference on Human Factors in Computing Systems\/},
  520--527, 1999.

\bibitem{chen2004}
Chen, J., Huang, Y.~A., dan Benesty, J.,
  {Time Delay Estimation},
  {\em {Audio Signal Processing for Next-Generation Multimedia Communication Systems}\/},
  Kluwer Academic Publishers, 2004.

\bibitem{cole1997}
Cole, R.~A., Mariani, J., Uszkoreit, H., Zaenen, A., dan Zue, V. (editor),
  {\em {Survey of State of the Art in Human Language Technology}\/},
  Cambrige University Press, 1997, \urlprefix\url{http://cslu.cse.ogi.edu/HLTsurvey/}.

\bibitem{forouzan2003}
Forouzan, B.~A.,
  {\em {Data Communications and Networking}\/}, edisi ke-3,
  McGraw-Hill, 2003.

\bibitem{foster2007}
Foster, M.,
  {Enhancing human-computer interaction with embodied conversational agents},
  {\em Universal Access in Human-Computer Interaction. Ambient Interaction\/},
  828--837, 2007.

\bibitem{gasperis2010}
Gasperis, G.~D.,
  {Building an AIML Chatter Bot Knowledge-Base Starting from a FAQ and a Glossary},
  {\em Journal of e-Learning and Knowledge Society\/}, \textbf{6}(2),
  75--83, 2010.

\bibitem{gelfand2010}
Gelfand, S.~A.,
  {\em {Hearing: An Introduction to Psychological and Physiological Acoustics}\/},
  edisi ke-5, Informa Healthcare, 2010.

\bibitem{goldstein2010}
Goldstein, E.~B.,
  {\em {Sensation and Perception}\/},
  edisi ke-8, Wadsworth, 2010.

\bibitem{halliday2004}
Halliday, D., Resnick, R., dan Walker, J.,
  {\em {Fundamentals of Physics}\/},
  edisi ke-7, Wiley, 2004.

\bibitem{hartmann2005}
Hartmann, B., Mancini, M., dan Pelachaud, C.,
  Implementing expressive gesture synthesis for embodied conversational agents,
  {\em Gesture Workshop\/}, 188--199, Springer, 2005. 

\bibitem{huang2001}
Huang, X., Acero, A., dan Hon, H.-W.,
  {\em {Spoken Language Processing: A Guide to Theory, Algorithm, and System Development}\/},
  edisi ke-1, Prentice Hall, 2001.

\bibitem{huang2008}
Huang, Y.~A., Benesty, J., dan Chen, J.,
  {Time Delay Estimation and Source Localization},
  {\em {Springer Handbook of Speech Processing}\/},
  Benesty, J., Sondhi, M.~M., dan Huang, Y.~A., editor,
  1043--1063, Springer, 2008.

\bibitem{jurafsky2009}
Jurafsky, D. dan Martin, J.~H.,
  {\em {Speech and Language Processing}\/},
  edisi ke-2, Prentice-Hall, 2009.

\bibitem{kollmeier2008}
Kollmeier, B., Brand, T., dan Meyer, B.,
  {Perception of Speech and Sound},
  {\em {Springer Handbook of Speech Processing}\/},
  Benesty, J., Sondhi, M.~M., dan Huang, Y.~A., editor, 
  61--82, Springer, 2008.

\bibitem{lee2005}
Lee, J., Ji, S., Hahn, M., dan Cho, Y.,
  {Real-Time Sound Localization Using Time Difference for Human-Robot Interaction},
  {\em Proceedings of 16th IFAC World Congress\/}, 2005.

\bibitem{lee2002}
Lee, S., Badler, J., dan Badler, N.,
  {Eyes alive},
  {\em ACM Transactions on Graphics (TOG)\/},
  \textbf{21}, 637--644, ACM, Juli 2002.

\bibitem{loebner2011}
Loebner, H.~G.,
  {\em Home Page of The Loebner Prize in Artificial Intelligence\/},
  \urlprefix\url{http://www.loebner.net/Prizef/loebner-prize.html},
  diakses pada 10 Mei 2011, 02:00 WIB.

\bibitem{martin2001}
Martin, A.~F. dan Przybocki, M.~A.,
  Speaker recognition in a multi-speaker environment,
  {\em EUROSPEECH-2001\/}, 787-790, 2001.

\bibitem{massaro2006}
Massaro, D., Liu, Y., Chen, T., dan Perfetti, C., 
  {A multilingual embodied conversational agent for tutoring speech and language learning},
  {\em Proceedings of the Ninth International Conference on Spoken Language Processing
  (Interspeech 2006-ICSLP)\/}, 825--828, 2006.

\bibitem{mctear2002}
McTear, M.~F.,
  {Spoken dialogue technology: enabling the conversational user interface,}
  {\em ACM Computing Surveys\/}, \textbf{34}, 90--169, Maret 2002.

\bibitem{morency2006}
Morency, L.-P., Christoudias, C.~M., dan Darrell, T.,
  {Recognizing gaze aversion gestures in embodied conversational discourse},
  {\em Proceedings of the 8th international conference on Multimodal interfaces\/},
  289--294, ACM, 2006.

\bibitem{nakano2010}
Nakano, A.~Y.,
  {\em Exploring {Spatial Information For Distant Speech Recognition Under
  Real Environmental Conditions}\/}, Disertasi Doktor,
  Toyohashi University of Technology, Maret 2010.

\bibitem{niewiadomski2009}
Niewiadomski, R., Bevacqua, E., Mancini, M., dan Pelachaud, C.,
  {Greta: an interactive expressive ECA system,}
  {\em Proceedings of The 8th International Conference on Autonomous Agents and
  Multiagent Systems - Volume 2\/}, 1399--1400,
  International Foundation for Autonomous Agents and Multiagent Systems, 2009.

\bibitem{nugraha2010}
Nugraha, A.~A., Taufiq, A., Utama, N.~I., dan Prihatmanto, A.~S.,
  {Smart Assistant for Museum's Objects Navigation (SAMsON)},
  {\em Proceedings of The 5th AOTULE International Postgraduate
  Students Conference on Engineering\/}, 186--189, November 2010.

\bibitem{park2009}
Park, Y.,
  {Application of 3D sound technology to intelligent robots},
  {\em Proceedings of the 3rd International Universal Communication Symposium
  on - IUCS '09\/}, 199--204, 2009.

\bibitem{rabiner2007}
Rabiner, L.~R. dan Schafer, R.~W.,
  {Introduction to digital speech processing,}
  {\em Foundations and Trends in Signal Processing\/}, \textbf{1}(1-2),
  January 2007.

\bibitem{rehm2005}
Rehm, M.,
  {From Chatterbots to Natural Interaction -- 
  Face to Face Communication with Embodied Conversational Agents},
  {\em IEICE Transactions on Information and Systems\/}, \textbf{E88-D}(11),
  2445--2452, November 2005.

\bibitem{rickel2002}
Rickel, J., Marsella, S., Gratch, J., Hill, R., Traum, D., dan Swartout, W.,
  {Toward a new generation of virtual humans for interactive experiences},
  {\em IEEE Intelligent Systems\/}, \textbf{17}(4), 32--38, Juli 2002.

\bibitem{shawar2007}
Shawar, B.~A. dan Atwell, E.,
  {Chatbots: Are they Really Useful?}
  {\em LDV Forum\/}, \textbf{22}(1), 29--49, 2007.

\bibitem{stephens-}
Stephens, K.~R.,
  {What Has the Loebner Contest Told Us About Conversant Systems?}
  \urlprefix\url{http://www.behavior.org/resources/325.pdf}, diakses
  pada 10 Mei 2011, 02:00 WIB.

\bibitem{swartout2010}
Swartout, W., et~al.,
  Ada and grace: toward realistic and engaging virtual museum guides,
  {\em Proceedings of the 10th international conference on Intelligent virtual agents (IVA'10)\/},
  286--300, Springer-Verlag, 2010.

\bibitem{turing1950}
Turing, A.~M.,
  {Computing Machinery and Intelligence}, {\em Mind\/},
  \textbf{59}(236), 433--460, Oktober 1950.

\bibitem{vanes2002}
{Van Es}, I., Heylen, D., {Van Dijk}, B., dan Nijholt, A.,
  {Making agents gaze naturally-Does it work?}
  {\em Proceedings of the Working Conference on Advanced Visual Interfaces\/},
  357--358, ACM, 2002.

\bibitem{wallace2003}
Wallace, R.~S.,
  {\em The Elements of AIML Style\/}, Maret 2003.

\bibitem{wallace2009}
Wallace, R.~S.,
  {The Anatomy of A.L.I.C.E.},
  {\em {Parsing the Turing Test}\/},
  Epstein, R., Roberts, G., dan Beber, G., editor,
  181--210, Springer, 2009.

\bibitem{weizenbaum1966}
Weizenbaum, J.,
  {ELIZA}-a computer program for the study of natural
  language comunication between man and machine,
  {\em Communications of the ACM\/}, \textbf{9}(1), 36--45, 1966.

\bibitem{alice2011}
\underline{\hspace{5em}}, {\em Chat Robot Survey\/}, {ALICE Artificial Intelligence Foundation},
  \urlprefix\url{http://www.alicebot.org/aimlbots.html}, diakses pada 10 Mei 2011, 02:00 WIB.

\bibitem{bps2010}
\underline{\hspace{5em}}, {\em Hasil Sensus Penduduk 2010: Data Agregat per Provinsi\/},
  {Badan Pusat Statistik}, 2010.

\bibitem{dppobappenas2009}
\underline{\hspace{5em}}, {\em Jumlah Pengunjung Museum di Indonesia\/},
  {Direktorat Kebudayaan, Pariwisata, Pemuda, dan Olahraga, Badan Perencanaan Pembangunan Nasional}, 2009,
  \urlprefix\url{http://kppo.bappenas.go.id/preview/225}, diakses pada 10 Mei 2011, 17:00 WIB.
  
\bibitem{themuseumofscienceboston-}
\underline{\hspace{5em}}, {\em Meet Ada and Grace: Virtual Museum Guides\/},
  {The Museum of Science, Boston}, \urlprefix\url{http://www.mos.org/interfaces/adagrace.php},
  diakses pada 10 Mei 2011, 17:00 WIB.

\bibitem{uscinstituteforcreativetechnologies-}
\underline{\hspace{5em}}, {\em Responsive Virtual Human Museum Guides\/},
  {USC Institute for Creative Technologies},
  \urlprefix\url{http://ict.usc.edu/projects/responsive_virtual_human_museum_guides/},
  diakses pada 10 Mei 2011, 17:00 WIB.

\bibitem{armedwithscience-}
\underline{\hspace{5em}}, {\em Virtual Humans Blog on Science and Women’s History\/},
  {Armed with Science}, Maret 2010,
  \urlprefix\url{http://science.dodlive.mil/2010/03/04/virtual-humans-blog-on-science-and-womens-history/},
  diakses pada 10 Mei 2011, 17:00 WIB.

\end{thebibliography}
